# Read data {#readdata}

> Prerequisite: Chapter 11 ‘Data Import’ from R for Data Science, available at http://r4ds.had.co.nz/data-import.html

## Introduction to Reading Data

Once you have downlaoded the data from the Understanding Society survey to your computer, the first thing you need to do is read the data into R. There are a number of ways to do this, and this class will cover loading data into R by using base R and both the readr and data.table packages.

The data we will be loading into R is the individual adult questionnaire answers from wave 1 (*UKDA-6614-tab/tab/us_w1/a_indresp.tab*). This should be saved in the *data* folder you created in your project folder (this was covered in the Introduction).

## Reading Data into R

### Using base R

To read this data into R using functions from base R, we can first use the **read.table** function. To read the data in correctly, we need to use *header = TRUE* as the first row of the data contains the names of the variables, and also *stringsAsFactors = FALSE*, which stops R from treating text variables as factors. We can convert these into factors later, when necessary.

```{r cache = TRUE}
UndSoc1 <- read.table("data/UKDA-6614-tab/tab/us_w1/a_indresp.tab",
                      header = TRUE,
                      stringsAsFactors = FALSE)
```

Different types of files need to read into R differently. The **read.table** function reads data that is *tab separated* (the file will be called 'name*.tab*'), the **read.csv** function reads data that is *comma separed* (the file will be called 'name*.csv*') and **read.delim** reads data that is separated in any way. These work in the same way, so once you've mastered one, you've mastered them all.

### Using the readr package

We can also read data into R using the package *readr*, part of the *tidyverse* package, with the **read_tsv** command. This is the equivalent command to **read.table** from base R.

```{r cache = TRUE}
# We can either load the entire 'tidyverse' package into R, which includes the readr package as well as others such as ggplot2, or just the 'readr' package on its own.

library(readr)

UndSoc2 <- read_tsv("data/UKDA-6614-tab/tab/us_w1/a_indresp.tab")
```

The readr package also allows us to read different types of files into R. The commands **read_csv** and **read_delim** read comma separated data and any data into R respectively.

### Using the data.table package

The **fread** function from the **data.table** package is probably the fastest way to read in the data. 

```{r cache = TRUE}
library(data.table)

system.time(UndSoc3 <- fread("data/UKDA-6614-tab/tab/us_w1/a_indresp.tab"))
```

It took less than 3 seconds!
 
With small data sets the choice between these three methods is not very important, but with larger data the gain in efficiency that **readr** and **data.table** provide can be quite substantial.

## Other data formats

In R you can easily read in data in other formats, such as csv files, Stata, SPSS, SAS, Excel and others. There are many tutorials on how to do this on the web. See, for example, https://www.datacamp.com/courses/importing-data-in-r-part-1/ and https://www.datacamp.com/courses/importing-data-in-r-part-2 .

## Saving the R workspace

Once you have read your data into R you can save it as R workspace.

``{r cache = TRUE}
# I will remove some objects from memory to speed things up
rm(UndSoc2, UndSoc3)
# saving R workspavce now in myData (you need to create myData first)
save.image("myData/readTest.RData")
```

Next time I need this file I can simply load the workspace.

``{r cache = TRUE}
# first let's remove everything from the workspace
rm(list = ls())
# load the workspace
system.time(load("myData/readTest.RData"))
```

Of course, in R workspace you can save not only data frames but any objects: models, plots, functions, etc.